\chapter{The SPARC Language}
\label{ch:sparc}

\begin{cluster}
\label{grp:prmbl:sparc::presents}

\begin{preamble}
\label{prmbl:sparc::presents}
This chapter presents \pml{}: a parallel and functional language used
throughout the book for specifying algorithms.

\pml is a ``strict'' functional language similar to the ML class of
languages such as Standard ML or SML, Caml, and F\#.
In pseudo code, we sometimes use mathematical notation, and even
English descriptions in addition to \pml{} syntax.
This chapter describes the basic syntax and semantics of \pml{}; we
introduce additional syntax as needed in the rest of the book.

\end{preamble}
\end{cluster}


\section{Syntax and Semantics of \pml{}}
\label{sec:sparc::syntax-and-semantics-of}

\begin{cluster}
\label{grp:grm:sparc::describes}

\begin{gram}
\label{grm:sparc::describes}
This section describes the syntax and the semantics of the core subset
of the \pml{} language.  The term~\defn{syntax}~refers to the
structure of the program itself, whereas the
term~\defn{semantics}~refers to what the program computes.
Since we wish to analyze the cost of algorithms, we are interested in
not just what algorithms compute, but how they compute.
Semantics that capture how algorithms compute are
called~\defn{operational semantics}, and when augmented with specific
costs,~\defn{cost semantics}.
Here we describe the syntax of \pml{} and present an informal
description of its operational semantics. 
We will cover the cost semantics of \pml in \chref{analysis::models}.
While we focus primarily on the core subset of \pml, we also describe
some~\defn{syntactic sugar}~that makes it easier to read or write code
without adding any real power.
Even though \pml is a strongly typed language, for our purposes in
this book, we use types primarily as a means of describing and
specifying the behavior of our algorithms.
We therefore do not present careful account of \pml's type system.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::shows}

\begin{gram}
\label{grm:sparc::shows}
The definition below shows the syntax of \pml{}.
A \pml{} program is an expression, whose syntax,
 describe the computations that can be expressed in \pml{}.
When evaluated an expression yield a value.
Informally speaking, evaluation of an expression proceeds involves
evaluating its sub-expressions to values and then combining these
values  to compute the value of the expression.
\pml{} is a strongly typed language, where every closed expression,
which have no undefined (free) variables, evaluates to a value or runs
forever.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:def:sparc::syntax}

\begin{definition}[\pml{} expressions]
\label{def:sparc::syntax}
\setlength{\tabcolsep}{20mm}
\begin{tabular}{llcll}
Identifier
& $id$ & := & $\ldots$
\\

Variables
& $x$ & := & $id$
\\

Type Constructors 
& $tycon$ & := & $id$
\\

Data Constructors 
& $dcon$ & := & $id$
\\

Patterns & $p$ 
& := & x & variable
\\
& & $\mid$  & $( p )$ & parenthesis
\\
& & $\mid$  & $p_1, p_2$ & pair
\\
& & $\mid$  & $dcon~( p )$ & data pattern
\\

Types & $\tau$ &  := & $\tyint$  & integers
\\
& &  $\mid$ &   $\tybool$  & booleans
\\
& &  $\mid$  & $\tau$ $[ * \tau ]^+$  & products
\\
& &  $\mid$  & $\tau \ra \tau$ & functions
\\
& & $\mid$ & $tycon$         &  type constructors
\\
& & $\mid$ & $dty$ & data types
\\

Data Types & $dty$
& := &   $dcon~[\cd{of}~\tau]$ 
\\
&  & $\mid$ &  $dcon~[ \cd{of}~\tau ]~\cd{|}~dty$
\\

Values & $v$
& := & $0  \mid  1  \mid  \ldots$ &  integers
\\
& & $\mid$ &  $-1  \mid -2  \mid \ldots$ &  integers
\\
& & $\mid$ & $\ctrue{} \mid  \cfalse{}$ & booleans
\\
& & $\mid$ & $\cd{not}  \mid \ldots$ & unary operations
\\
& & $\mid$ & $\cand  \mid  \cd{plus}  \mid  \ldots$ & binary operations
\\
& & $\mid$ & $v_1, v_2$  & pairs
\\
& & $\mid$ & $( v )$   & parenthesis
\\
& & $\mid$ & $dcon~( v )$  & constructed data
\\
& & $\mid$ & $\cfn{p}{e}$ & lambda functions
\\

Expression  & $e$ 
& := & $x$ & variables
\\
& & $\mid$ & $v$ & values
\\
& & $\mid$ & $e_1$ op $e_2$ & infix operations
\\
& & $\mid$ & $e_1, e_2$ & sequential pair
\\
& & $\mid$ & $e_1 \cd{||} e_2$ & parallel pair
\\
& & $\mid$ & $( e )$ & parenthesis
\\
& & $\mid$ & $\ccase~e_1~[\cd{|}~p~\cdra~e_2]^+$  & case
\\
& & $\mid$ & $\cif~e_1~\cthen~e_2~\celse~e_3$  & conditionals
\\
& & $\mid$ & $e_1~e_2$ & \mbox{function application}
\\
& & $\mid$ & $\clet~b^+~\cin~e~\cend$ & \mbox{local bindings}
\\

Operations & $op$ & := & $+ \mid - \mid * \mid - \ldots$
\\

Bindings &  $b$ & := & 
$x ( p )$ = $e$ & bind function
\\
&  & $\mid$ & $p = e$ & bind pattern
\\
&  & $\mid$ & $\ctype~tycon = \tau$ & bind type 
\\
&  & $\mid$ & $\ctype~tycon = dty$ & bind datatype
\\

\end{tabular}

\end{definition}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::identifiers}

\begin{gram}[Identifiers]
\label{grm:sparc::identifiers}
In \pml, variables, type constructors, and data constructors are given
a name, or an~\defn{identifier}.  
An identifier consist of only alphabetic and numeric characters (a-z,
A-Z, 0-9), the underscore character (``\_''), and optionally end with
some number of ``primes''.  
Example identifiers include, $x'$, $x_1$, $x_l$, $\cdvar{myVar}$,
$\cdvar{myType}$, $\cdvar{myData}$, and $\cdvar{my\_data}$.



Program~\defn{variables},~\defn{type constructors}, and~\defn{data
  constructors}~are all instances of identifiers.
During evaluation of a \pml expression, variables are bound to values,
which may then be used in  a computation later.  
In \pml, variable are~\defn{bound}~during function application, as part
of matching the formal arguments to a function to those specified by
the application, and also by $\cd{let}$ expressions.
If, however, a variable appears in an expression but it is not bound
by the expression, then it is~\defn{free}~in the expression.
We say that an expression is~\defn{closed}~if it
has no free variables.


Types constructors give names to types.  For example, the type of
binary trees may be given the type constructor $\cdvar{btree}$.
Since for the purposes of simplicity, we rely on mathematical rather
than formal specifications, we usually name our types behind
mathematical conventions.
For example, we denote the type of natural numbers by $\tynat$, the type
of integers by $\tyint$, and the type of booleans by $\tybool$.


Data constructors serve the purpose of making complex data structures.
By convention, we will capitalize data constructors, while starting
variables always with lowercase letters.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::patterns}

\begin{gram}[Patterns]
\label{grm:sparc::patterns}
In \pml, variables and data constructors can be used to construct more
complex~\defn{patterns}~over data.
For example, a pattern can be a pair $(x,y)$, or a triple of
variables $(x,y,z)$, or it can consist of a data constructor
followed by a pattern, e.g., $\cdvar{Cons}(x)$ or $\cdvar{Cons}(x,y).$
Patterns thus enable a convenient and concise way to pattern match
over the data structures in \pml.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::built-in-types}

\begin{gram}[Built-in Types]
\label{grm:sparc::built-in-types}
Types of \pml include base types such as integers $\tyint$, booleans
$\tybool$, product types such as $\tau_1 * \tau_2 \ldots \tau_n$,
function types $\tau_1 \ra \tau_2$ with domain $\tau_1$ and range
$\tau_2$, as well as user defined data types.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::data-types}

\begin{gram}[Data Types]
\label{grm:sparc::data-types}
In addition to built-in types, a program can define new~\defn{data types}~as a
union of tagged types, also called variants, by ``unioning'' them via
distinct~\defn{data constructors}.
For example, the following data type defines a point as a
two-dimensional or a three-dimensional coordinate of integers.
\[
\begin{array}{lcl}
\cd{type}~\cdvar{point} & = & \cdvar{PointTwo}~\cd{of}~\tyint * \tyint
\\
           & | & \cdvar{Point3D}~\cd{of}~\tyint * \tyint * \tyint
\end{array} 
\]

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::recursive-data-types}

\begin{gram}[Recursive Data Types]
\label{grm:sparc::recursive-data-types}
In \pml recursive data types are relatively easy to define and compute
with. For example, we can define a point list data type as follows
\[
\begin{array}{l}
\cd{type}~\cdvar{plist} = \cdvar{Nil}~|~\cdvar{Cons}~\cd{of}~\cdvar{point} * \cdvar{plist}.
\end{array}
\] 
Based on this definition the list 
\[
\begin{array}{l}
\cdvar{Cons}(\cdvar{PointTwo} (0,0),  
\\
~~~~~~~~~~\cdvar{Cons}(\cdvar{PointTwo} (0,1), 
\\
~~~~~~~~~~~~~~~~~~~~\cdvar{Cons}(\cdvar{PointTwo}(0,2), \cdvar{Nil})))  
\end{array} 
\]
defines a list consisting of three points.

\end{gram}
\end{cluster}

\begin{flex}
\label{grp:xrcs:sparc::booleans}

\begin{exercise}[Booleans]
\label{xrcs:sparc::booleans}
Some built-in types such as booleans, $\tybool$, are in fact syntactic
sugar and can be defined by using union types as follows.
Describe how you can define booleans using data types of \pml{}.

\end{exercise}

\begin{solution}
\label{sol:sparc::booleans}
Booleans can be defined as follows.

\[
\begin{array}{l}
\cd{type}~\cdvar{myBool} = \cdvar{myTrue}~|~\cdvar{myFalse} 
\end{array}
\] 

\end{solution}
\end{flex}

\begin{cluster}
\label{grp:grm:sparc::option-type}

\begin{gram}[Option Type]
\label{grm:sparc::option-type}
Throughout the book, we use~\defn{option}~types quite frequently.
Option types for natural numbers can be defined as follows.
\[
\begin{array}{l}
\cd{type}~\cdvar{option} = \cdvar{None} ~|~ \cdvar{Some}~\cd{of}~\tynat
\end{array}
\]
Similarly, we can define option types for integers.
\[
\begin{array}{l}
\cd{type}~\cdvar{intOption} = \cdvar{INone} ~|~ \cdvar{ISome}~\cd{of}~\tyint
\end{array}
\]
Note that we used a different data constructor for naturals.  
This is necessary for type inference and type checking.
Since, however, types are secondary for our purposes in this book, we
are sometimes sloppy in our use of types for the sake of simplicity.
For example, we use throughout $\cdvar{None}$ and $\cdvar{Some}$ for option
types regardless of the type of the contents.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:tch:sparc::todo}

\begin{teachnote}
\label{tch:sparc::todo}
TODO: SEQUENCES etc.SETS.

\end{teachnote}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::values}

\begin{gram}[Values]
\label{grm:sparc::values}
Values of \pml, which are the irreducible units of computation
include natural numbers, integers, Boolean values $\ctrue$ and $\cfalse$,
unary primitive operations, such as boolean negation $\cd{not}$,
arithmetic negation $\cminus$, as well as binary operations such as
logical and $\cand$ and arithmetic operations such as $\cplus$.
Values also include constant-length tuples, which correspond to
product types, whose components are values.
Example tuples used commonly through the book include binary tuples or
pairs, and ternary tuples or triples.
Similarly, data constructors applied to values, which correspond to
sum types, are also values.

As a functional language, \pml treats all function as values. 
The anonymous function
$\cd{lambda}~p.~e$
is a function whose arguments are specified by the pattern $p$, and
whose body is the expression $e$.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:xmpl:sparc::function}

\begin{example}
\label{xmpl:sparc::function}
\begin{itemize}

\item

The function 
$\cd{lambda}~x. x + 1$ takes a single variable as an argument and
adds one to it.

\item
The function 
$\cd{lambda}~(x,y).~x$ takes a pairs as an argument and
returns the first component of the pair.
\end{itemize}

\end{example}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::expressions}

\begin{gram}[Expressions]
\label{grm:sparc::expressions}
Expressions, denoted by $e$ and variants (with subscript, superscript,
prime), are defined inductively, because in many cases, an expression
contains other expressions.
Expressions describe the computations that can be expressed in \pml. 
Evaluating an expression via the operational semantics of \pml
produce the value for that expression.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::infix-expressions}

\begin{gram}[Infix Expressions]
\label{grm:sparc::infix-expressions}
An~\defn{infix expression}, $e_1~\cdvar{op}~e_2$, involve two expressions
and an infix operator $\cdvar{op}$.  The infix operators include $+$
(plus), $-$ (minus), $*$ (multiply), $/$ (divide),
$<$ (less), $>$ (greater), $\cd{or}$, and $\cd{and}.$
For all these operators the infix expression
$e_1~\cdvar{op}~e_2$ is just syntactic sugar for $f(e_1, e_2)$ where
$f$ is the function corresponding to the operator $\cdvar{op}$ (see
parenthesized names that follow each operator above).  

We use standard precedence rules on the operators to indicate their
parsing.  For example in the expression
\[
\cd{3 + 4 * 5}
\]
the $*$ has a higher precedence than $+$ and therefore the
expression is equivalent to $3 + (4 * 5)$. 

Furthermore all operators are left associative unless stated
otherwise, i.e., that is to say that $a~\cdvar{op}_1~b~\cdvar{op}_2~c =
(a~\cdvar{op}_1~b)~\cdvar{op}_2~c$ if $\cdvar{op}_1$ and $\cdvar{op}_2$ have the
same precedence.  

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:xmpl:sparc::expressions}

\begin{example}
\label{xmpl:sparc::expressions}
The expressions $5 - 4 + 2$ evaluates to $(5-4) + 2 = 3$ not
$5 - (4 + 2) = -1$, because $-$ and $+$ have the same
precedence.

\end{example}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::sequential-and-parallel-composition}

\begin{gram}[Sequential and Parallel Composition]
\label{grm:sparc::sequential-and-parallel-composition}
Expressions include two special infix operators: ``$,$'' and
$||$, for generating ordered pairs, or tuples, either
sequentially or in parallel.

The~\defn{comma} operator or~\defn{sequential composition}~as in the
infix expression $(e_1, e_2)$, evaluates $e_1$ and
$e_2$ sequentially, one after the other, and returns the ordered pair
consisting of the two resulting values.
Parenthesis delimit  tuples.

The~\defn{parallel} operator or~\defn{parallel composition}~``$||$'',
as in the infix expression
$(e_1~||~e_2)$,
evaluates $e_1$ and $e_2$ in parallel, at the same time, and
returns the ordered pair consisting of the two resulting values.


The two operators are identical in terms of their return values.
However, we will see later, their cost semantics differ: one is
sequential and the other parallel.  The comma and parallel operators
have the weakest, and equal, precedence.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:tch:sparc::cost}

\begin{teachnote}
\label{tch:sparc::cost}
Cost model pointer.

\end{teachnote}
\end{cluster}

\begin{cluster}
\label{grp:xmpl:sparc::expression}

\begin{example}
\label{xmpl:sparc::expression}
\begin{itemize}

\item

The expression 
\[
\begin{array}{l}
\cd{lambda}~(x, y).~(x * x, y * y)
\end{array}
\]
is a function that take two arguments $x$ and $y$ and returns a
pair consisting of the squares $x$ and $y$.

\item
The expression
\[
\begin{array}{l}
\cd{lambda}~(x, y).~(x * x~||~y * y)
\end{array}
\]
is a function that take two arguments $\cd{x}$ and $\cd{y}$ and returns a
pair consisting of the squares $\cd{x}$ and $\cd{y}$ by squaring each of
$\cd{x}$ and $\cd{y}$ in parallel.
\end{itemize}

\end{example}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::case-expressions}

\begin{gram}[Case Expressions]
\label{grm:sparc::case-expressions}
A~\defn{case expression}~such as 
\[
\begin{array}{l}
\cd{case}~e_1 \\
\cd{| Nil}\dra e_2 \\ 
\cd{| Cons}~(x,y)\dra e_3 \\
\end{array}
\]
first evaluates the expression $e_1$ to a value $v_1$, which must
return data type.
It then matches $v_1$ to one of the patterns, $\cdvar{Nil}$ or
$\cdvar{Cons}~(x,y)$ in our example, binds the variable if any in the
pattern to the respective sub-values of $v_1$, and evaluates the
``right hand side'' of the matched pattern, i.e., the expression $e_2$
or $e_3$.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::conditionals}

\begin{gram}[Conditionals]
\label{grm:sparc::conditionals}
A conditional or an~\defn{if-then-else expression},
$\cd{if}~e_1~\cd{then}~e_2~\cd{else}~e_3$, evaluates the expression $e_1$,
which must return a Boolean.
If the value of $e_1$ is true then the result of the if-then-else
expression is the result of evaluating $e_2$, otherwise it is the
result of evaluating $e_3$.  
This allows for conditional evaluation of expressions.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::function-application}

\begin{gram}[Function Application]
\label{grm:sparc::function-application}
A~\defn{function application}, $e_1~e_2$, applies the function
generated by evaluating~$e_1$ to the value generated by
evaluating~$e_2$.  
For example, lets say that~$e_1$ evaluates to the function~$f$
and~$e_2$ evaluates to the value~$v$, then we apply~$f$ to~$v$ by
first matching~$v$ to the argument of $f$, which is pattern, to
determine the values of each variable in the pattern.
We then substitute in the body of~$f$ the value of each variable for
the variable.  To~\defn{substitute}~a value in place of a variable~$x$
in an expression~$e$, we replace each instance of~$x$ with~$v$.

For example if function $\cd{lambda}~(x,y).~e$ is applied to the pair
$\cd{(2,3)}$ then $x$ is given value $\cd{2}$ and $y$ is given value
$\cd{3}$. 
Any free occurrences of the variables $x$ and $y$ in the
expression $e$ will now be bound to the values $\cd{2}$ and $\cd{3}$
respectively.  
We can think of function application as substituting
the argument (or its parts) into the free occurrences of the variables
in its body $e$.
The treatment of function application is why we call \pml{} a~\defn{strict}~language. 
In strict or call-by-value languages, the argument to the function is
always evaluated to a value before applying the function.
In contrast non-strict languages wait to see if the argument will be
used before evaluating it to a value. 

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:xmpl:sparc::lambda}

\begin{example}
\label{xmpl:sparc::lambda}
\begin{itemize}

\item
The expression
\[
(\cd{lambda}~(x,y).~x / y)~(8,2)
\]
\\
evaluates to $4$ since $8$ and 
$2$ are bound to $x$ and $y$, respectively, and then divided.

\item
The expression 
\[
(\cd{lambda}~(f,x).~f(x,x))~(\cdvar{plus},3)
\] 
evaluates to $6$ because $f$ is bound to the function
$\cdvar{plus}$, $x$ is bound to $3$, and then $\cdvar{plus}$ is applied
to the pair $(3,3)$.

\item
The expression
\[
(\cd{lambda}~x.~(\cd{lambda}~y .~x + y ) )~3
\]
\\
evaluates to a function that adds $3$ to any integer.
\end{itemize}

\end{example}
\end{cluster}

\begin{cluster}
\label{grp:grm:sparc::bindings}

\begin{gram}[Bindings]
\label{grm:sparc::bindings}
The~\defn{let expression}, 
\[
\cd{let}~b^+\cd{in}~e~\cd{end},
\] 
consists of a sequence of bindings $b^+$, which define local variables
and types, followed by an expression $e$, in which those bindings are
visible.  In the syntax for the bindings, the superscript $+$ means
that $b$ is repeated one or more times.  Each binding $b$ is either a
variable binding, a function binding, or a type binding.
The let expression evaluates to the result of evaluating $e$ given the
variable bindings defined in $b$.

A~\defn{function binding}, $x (p) = e$, consists of a function
name, $x$ (technically a variable), the arguments for the function,
$p$, which are themselves a pattern, and the body of the function,
$e$. 


Each~\defn{type binding}~equates a type to a base type or a data type.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:xmpl:sparc::consider}

\begin{example}
\label{xmpl:sparc::consider}
Consider the following let expression.
\[
\begin{array}{l}
\cd{let}\\ 
~~~~x = 2 + 3\\
~~~~f (w) = (w * 4, w - 2)\\
~~~~(y,z) = f(x-1)\\
\cd{in}\\ 
~~~~x + y + z\\
\cd{end} 
\end{array}
\]

The first  binding the variable $x$ to $\cd{2 + 3 = 5}$;
The second binding defines a function $f(w)$ which returns a pair;
The third binding applies the function $f$ to $x - 1 = 4$
returning the pair $(4 * 4, 4 -2) = (16, 2)$, which
  $y$ and $z$ are bound to, respectively (i.e., $y = 16$ and
  $z = 2$.
Finally the let expressions adds $x, y, z$ and yields $5 + 16 + 2$.
The result of the expression is therefore $23$.

\end{example}
\end{cluster}

\begin{cluster}
\label{grp:nt:sparc::careful}

\begin{note}
\label{nt:sparc::careful}
Be careful about defining which variables each binding can see, as
this is important in being able to define recursive functions.  In
\pml{} the expression on the right of each binding in a $\cd{let}$ can
see all the variables defined in previous variable bindings, and can
see the function name variables of all binding (including itself)
within the $\cd{let}$.
Therefore the function binding
\[
\cd{x}(p) = e
\]
is not equivalent to the variable binding
\[
\cd{x} = \cd{lambda}~p.e,
\]
because in the prior $x$ can be used
in $e$ and in the later it cannot.
Function bindings therefore allow for the definition of
recursive functions.  
Indeed they allow for mutually recursive functions since the body of
function bindings within the same $\cd{let}$ can reference each other.

\end{note}
\end{cluster}

\begin{cluster}
\label{grp:xmpl:sparc::else}

\begin{example}
\label{xmpl:sparc::else}
The expression
\[
\begin{array}{l}
\cd{let}\\
~~~~f(i) = \cd{if} ~(i < 2) ~\cd{then}~ i ~\cd{else}~ i  * 
f(i - 1) \\
\cd{in} \\ 
~~~~f(5) \\
\cd{end} 
\end{array}
\]
will evaluate to the factorial of $5$, i.e., $5 * 4 * 3 * 2
* 1$, which is $120$.

\end{example}
\end{cluster}

\begin{cluster}
\label{grp:xmpl:sparc::piece}

\begin{example}
\label{xmpl:sparc::piece}
The piece of code below illustrates an example use of data types and
higher-order functions.
\[
\begin{array}{l}
\cd{let}
\\ 
~~~~\cd{type}~\cdvar{point} = \cdvar{PointTwo}~\cd{of}~\tyint * \tyint
\\ 
~~~~~~~~~~~~~~~~~~~~~~~~|~~\cdvar{PointThree}~\cd{of}~ \tyint * \tyint * \tyint
\\
~~~~\cdvar{injectThree}~(\cdvar{PointTwo}~(x, y)) = \cdvar{PointThree}~(x, y, 0)
\\  
~~~~\cdvar{projectTwo}~(\cdvar{PointThree}~(x, y, z)) = \cdvar{PointTwo}~(x, y)
\\ 
~~~~\cdvar{compose}~f~g = f~g  
\\
~~~~p0 = \cdvar{PointTwo}~(0,0)
\\
~~~~q0 = \cdvar{injectThree}~p0
\\
~~~~p1 = (\cdvar{compose}~\cdvar{projectTwo}~\cdvar{injectThree})~p0
\\
\cd{in} 
\\
~~~~(p0, q0)
\\
\cd{end}
\end{array}
\]

The example code above defines a $\cdvar{point}$ as a two (consisting of
$x$ and $y$ axes) or three dimensional (consisting of $x$, $y$, and
$z$ axes) point in space.
The function $\cdvar{injectThree}$ takes a 2D point and transforms it to a 3D
point by mapping it to a point on the $z=0$ plane. 
The function $\cdvar{projectTwo}$ takes a 3D point and transforms it to a 2D
point by dropping its $z$ coordinate.
The function $\cdvar{compose}$ takes two functions $f$ and $g$ and composes
them.
The function $\cdvar{compose}$ is a higher-order function, since id operates
on functions.

The point $p0$ is the origin in 2D.  The point $q0$ is then computed
as the origin in 3D.  The point $p1$ is computed by injecting $p0$ to
3D and then projecting it back to 2D by dropping the $z$ components,
which yields again $p0$.  
In the end we thus have $p0 = p1 = (0,0)$. 

\end{example}
\end{cluster}

\begin{cluster}
\label{grp:xmpl:sparc::defines}

\begin{example}
\label{xmpl:sparc::defines}
The following \pml code, which defines a binary tree whose leaves and
internal nodes holds keys of integer type.
The function $\cdvar{find}$ performs a lookup in a given binary-search tree
$t$, by recursively comparing the key $x$ to the keys along a path in
the tree.

\[
\begin{array}{l}
\cd{type}~\cdvar{tree} = \cdvar{Leaf}~\cd{of}~\tyint~|~\cdvar{Node}~\cd{of}~(\cdvar{tree}, \tyint, \cdvar{tree})
\\
\cdvar{find}~(t, x) = 
\\
~~~~\cd{case}~t
\\ 
~~~~|~\cdvar{Leaf}~y \dra x = y 
\\
~~~~|~\cdvar{Node}~(\cdvar{left}, y, \cdvar{right}) \dra
\\
~~~~~~~~~\cd{if}~x = y~\cd{then} 
\\
~~~~~~~~~~~~~\cd{return}~\cd{true}
\\
~~~~~~~~~\cd{else}~\cd{if}~x < y~\cd{then} 
\\
~~~~~~~~~~~~~\cdvar{find}~(\cdvar{left}, x)
\\
~~~~~~~~~\cd{else}
\\
~~~~~~~~~~~~~\cdvar{find}~(\cdvar{right}, x)
\end{array}
\]

\end{example}
\end{cluster}

\begin{cluster}
\label{grp:rmrk:sparc::takes}

\begin{remark}
\label{rmrk:sparc::takes}
~\\
The definition
\[
\cfn{x}{(\cfn{y}{f(x,y)})}
\]
  takes a function $f$ of a pair of arguments and converts it
  into a function that takes one of the arguments and returns a
  function which takes the second argument.  This technique can be
  generalized to functions with multiple arguments and is often
  referred to as~\defn{currying}, named after Haskell Curry
  (1900-1982), who developed the idea.  It has nothing to do with the
  popular dish from Southern Asia, although that might be an easy way
  to remember the term.

\end{remark}
\end{cluster}

