\chapter{Sets and Relations}
\label{ch:bg::sets-and-relations}

\begin{cluster}
\label{grp:prmbl:bg::sets-and-relations::presents}

\begin{preamble}
\label{prmbl:bg::sets-and-relations::presents}
This chapter presents a review of some basic definitions on sets and
relations.

\end{preamble}
\end{cluster}


\section{Sets}
\label{sec:bg::sets}

\begin{cluster}
\label{grp:grm:bg::sets-and-relations::collection}

\begin{gram}
\label{grm:bg::sets-and-relations::collection}
A~\defn{set} is a collection of distinct objects.  The objects that
are contained in a set, are called~\defn{members} or
the~\defn{elements} of the set.  The elements of a set must be
distinct: a set may not contain the same element more than once. The
set that contains no elements is called the~\defn{empty set} and is
denoted by $\{\}$ or $\emptyset$.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:grm:bg::sets-and-relations::specification-of-sets}

\begin{gram}[Specification of Sets]
\label{grm:bg::sets-and-relations::specification-of-sets}
Sets can be specified intentionally, by mathematically describing
their members.
For example, the set of natural numbers, traditionally written as
$\tynat$, can be specified~\defn{intentionally} as the set of all
nonnegative integral numbers.
Sets can also be specified~\defn{extensionally} by listing their members.
For example, the set $\tynat = \{0, 1, 2, \ldots\}.$
We say that an element $x$ is a {\em member of} $A$, written $x \in
A$, if $x$ is in $A$.  
More generally, sets can be specified using~\defn{set comprehensions},
which offer a compact and precise way to define sets by mixing
intentional and extensional notation.

\end{gram}
\end{cluster}

\begin{cluster}
\label{grp:def:bg::sets-and-relations::union-and-intersection}

\begin{definition}[Union and Intersection]
\label{def:bg::sets-and-relations::union-and-intersection}
For two sets $A$ and $B$, the~\defn{union} $A \cup B$ is defined as
the set containing all the elements of $A$ and $B$.  Symmetrically,
their~\defn{intersection}, $A \cap B$ is the defined as the set
containing the elements that are member of both $A$ and $B$.
We say that $A$ and $B$ are~\defn{disjoint} if their intersection is
the empty set, i.e., $A \cap B = \emptyset$.

\end{definition}
\end{cluster}

\begin{flex}
\label{grp:def:bg::sets-and-relations::cartesian-product}

\begin{definition}[Cartesian Product]
\label{def:bg::sets-and-relations::cartesian-product}
Consider two sets~$A$ and~$B$.  The~\defn{Cartesian product $A \times
  B$} is the set of all ordered pairs $(a,b)$ where $a \in A$ and $b
\in B$, i.e.,
\[
A \times B = \cset{(a,b) : a \in A, b \in B}.
\]

\end{definition}

\begin{example}
\label{ex:bg::sets::cartesian}
The Cartesian product of $A = \cset{0,1,2,3}$ and $B = \cset{a,b}$ is
\[
\begin{array}{lll}
A \times B = & \{ & (0,a),(0,b),(1,a),(1,b),
\\
             &    & (2,a),(2,b),(3,a),(3,b) 
\\
             & \}.
\end{array}
\]

\end{example}
\end{flex}

\begin{flex}
\label{grp:def:bg::sets-and-relations::set-partition}

\begin{definition}[Set Partition]
\label{def:bg::sets-and-relations::set-partition}
Given a set $A$, a partition of $A$ is a set $P$ of non-empty subsets
of $A$ such that each element of $A$ is in exactly one subset in $P$.
We refer to each element of $P$ as a~\defn{block} or a~\defn{part} and
the set $P$ as a~\defn{partition} of $A$.
More precisely, $P$ is a partition of $A$ if the following conditions
hold:
\begin{itemize}
\item if $B \in P$, then $B \not= \emptyset$,
\item if $A = \bigcup_{B \in P}{B}$, and
\item if $B, C \in P$, then $B = C$ or $B \cap C = \emptyset$.
\end{itemize}

\end{definition}

\begin{example}
\label{xmpl:bg::sets-and-relations::contained}
If $A = \{1, 2, 3, 4, 5, 6 \}$ then $P = \{ \{1,3,5\}, \{2, 4, 6\} \}$
is a partition of $A$.  The set $\{1,3,5\}$ is a block.

The set $Q = \{ \{1,3,5,6\}, \{2, 4, 6\} \}$ is a not partition of
$A$, because the element $6$ is contained multiple blocks.

\end{example}
\end{flex}

\begin{flex}
\label{grp:def:bg::sets-and-relations::kleene-operators}

\begin{definition}[Kleene Operators]
\label{def:bg::sets-and-relations::kleene-operators}
For any set $\Sigma$, its~\defn{Kleene star} $\Sigma^*$ is the set of
all possible strings consisting of members of $\Sigma$, including the
empty string.

For any set $\Sigma$, its~\defn{Kleene plus} $\Sigma^+$ is the set of
all possible strings consisting of members $\Sigma$, excluding the
empty string.

\end{definition}

\begin{example}
\label{xmpl:bg::sets-and-relations::given}
Given $\Sigma = \{\texttt{a},\texttt{b}\}$,
\[
\begin{array}{rl}
\Sigma^* 
= 
\{
&
\str{}, 
\\
&
\str{\texttt{a}}, \str{\texttt{b}}, 
\\
&
\str{\texttt{aa}}, \str{\texttt{ab}}, 
\str{\texttt{ba}}, \str{\texttt{bb}}, 
\\
&
\str{\texttt{aaa}}, \str{\texttt{aab}}, \str{\texttt{aba}},
\str{\texttt{abb}},
\\
& \str{\texttt{baa}}, \str{\texttt{bab}}, \str{\texttt{bba}}, \str{\texttt{bbb}},
\\
&
\ldots
\\
\} &
\end{array}
\]
and
\[
\begin{array}{rl}
\Sigma^+ 
= 
\{
&
\str{\texttt{a}}, \str{\texttt{b}}, 
\\
&
\str{\texttt{aa}}, \str{\texttt{ab}}, 
\str{\texttt{ba}}, \str{\texttt{bb}}, 
\\
&
\str{\texttt{aaa}}, \str{\texttt{aab}}, \str{\texttt{aba}}, \str{\texttt{abb}}, 
\\
&
\str{\texttt{baa}}, \str{\texttt{bab}}, \str{\texttt{bba}}, 
\str{\texttt{bbb}},
\\
& \ldots
\\
\} & 
\\
\end{array}
\]

\end{example}
\end{flex}

\begin{cluster}
\label{grp:xrcs:bg::sets-and-relations::prove}

\begin{exercise}
\label{xrcs:bg::sets-and-relations::prove}
Prove that Kleene star and Kleene plus are closed under
string concatenation. 

\end{exercise}
\end{cluster}


\section{Relations}
\label{sec:bg::relations}

\begin{cluster}
\label{grp:def:bg::sets-and-relations::relation}

\begin{definition}[Relation]
\label{def:bg::sets-and-relations::relation}
A~\defn{(binary)~\defn{relation} from a set~$A$ to set~$B$} is a
subset of the Cartesian product of~$A$ and~$B$.  
For a relation~$R \subseteq A \times B$, 
\begin{itemize}
\item the set~$\cset{a : (a,b) \in
  R}$ is referred to as the~\defn{domain} of $R$, and 
\item 
the set~$\cset{b
  : (a,b) \in R}$ is referred to as the~\defn{range} of $R$.
\end{itemize} 

\end{definition}
\end{cluster}

\begin{cluster}
\label{grp:def:bg::sets-and-relations::function}

\begin{definition}[Function]
\label{def:bg::sets-and-relations::function}
A~\defn{mapping} or
\defn{function from $A$ to $B$}
is a relation $R \subset A \times B$
such that $|R| = |\mbox{domain}(R)|$, i.e., for every $a$ in the
domain of $R$ there is only one $b$ such that $(a,b) \in R$.  
The \defn{range} of the function is the range of $R$.
We call the set $A$ the \defn{domain} and the $B$ the \defn{co-domain}
of the function.

\end{definition}
\end{cluster}

\begin{cluster}
\label{grp:ex:bg::relations::sequencesdef}

\begin{example}
\label{ex:bg::relations::sequencesdef}
Consider the sets $A = \cset{0,1,2,3}$ and $B = \cset{a,b}$.

The set:
\[X = \cset{(0,a),(0,b),(1,b),(3,a)}\]
is a relation from $A$ to $B$ since $X \subset A \times B$, but not a mapping (function) since $0$
is repeated.   

The set
\[Y = \cset{(0,a),(1,b),(3,a)}\]
is both a relation and a function from $A$ to $B$ since each element
only appears once on the left.  

\end{example}
\end{cluster}

