\titledquestion{True/False}
\begin{parts}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\trueorfalse 

Consider taking the union of two sets with $n$ elements in total
(summed over both sets).  In the worst case, the union operation
performs the most work when the two sets contain an equal number of
elements and less work when one set contains significantly more
elements that the other.

\atrue

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\trueorfalse

On an undirected graph if  BFS visits a vertex $v$ at round $r$ (round
start at $0$) then the shortest path from $v$ to the source vertex has
length $r$.

\atrue


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\trueorfalse

When BFS visits a vertex $v$, it adds all of its neighbors to the
frontier, which is to be visited in the next round.

\afalse

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\trueorfalse

BFS is a natural algorithm for computing topological sort in a directed graph.

\afalse



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\trueorfalse
A back-edge is an edge of the DFS-tree that connects a descendant to
an ancestor.
\afalse
% a back-edge is a non-tree edge.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\trueorfalse 

In a DFS of a graph, for any edge $(u,v)$ in the graph, the discovery
time of $u$ is less than the discovery time of $v$.  

\afalse


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\trueorfalse 
For any DFS-tree edge $(u,v)$ in a graph, the finish time of $v$ is
less than the finish time of $u$.  
\atrue

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\trueorfalse 
Given a graph and the DFS numbers, we can determine the
forward, back, and cross edges by inspecting DFS numbers only.  
\atrue

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\trueorfalse

If a graph has negative edge weights, the unmodified Dijkstra's
algorithm for shortest path may loop forever.

\afalse

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\trueorfalse

If we run Bellman-Ford on a graph with a negative-weight cycle, then
the algorithm will terminate and indicate that it has found a
negative-weight cycle.

\afalse
% It can be unreachable from the source.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\trueorfalse 

A round of graph contraction with star partitioning (star contraction)
eliminates a constant fraction of vertices and edges.

\afalse

% not that of edges.


\end{parts}
