\section{Problems}

\begin{probl}{Sub-paths}
Prove that the sub-paths property holds for any graph, also in the
presence of negative weights.
\end{probl}

\begin{probl}{Distances and Paths}
We mentioned that we are sometimes only interested in the weight of a
shortest path, rather than the path itself.
%
For a weighted graph $G = (V,E,w)$, assume you are given the distances
$\dist_G(s,v)$ for all vertices $v \in V$.  For a particular vertex
$u$, describe how you could determine the vertices $P$ in a shortest
path from $s$ to $u$ in $O(p)$ work, where $p = \sum_{v \in P}
d^-(v)$.
\end{probl}

\begin{probl}{Dijkstra and BFS}
Consider a graph $G$ where all edges have weight $1$. The following is
true: When run on $G$, Dijkstra's can algorithm visits the vertices in
the same order as BFS does.
%
Explain your answer.
\end{probl}

\begin{probl}{Parallel Dijkstra}
As we have seen, there is not much parallelism in Dijkstra's
algorithm.
%
One way to increase parallelism in Dijkstra is to remove all vertices
that have the same priority together and visit them all at once.
%
Carefully describe an algorithm that can perform such visits in
parallel and the priority-queue ADT and the data structure needed by
the algorithm. 
%
What is the work and span of your algorithm? 
\end{probl}

\begin{probl}{Dijkstra with Negative Edge Weights}
Given a graph $G$ with possibly negative edge weights, one idea would
be to find the smallest negative edge weight $k$ and create a new
graph $G'$ by adding to each edge weight the value $|k|$ to ensure
that all edge weights are non-negative.  We can now run Dijkstra's
algorithm on $G'$ from any given source $s$ to find the shortest paths
to all the other vertices from $s$.  Since by adding the same constant
value $|k|$ to each edge, we have not changed the relative comparison
between edges.  We might thus conclude that the shortest paths in $G'$
and $G$ are the same.  Is this correct? Prove or disprove by creating
an example.  Is there a relationship between shortest paths in $G$ and $G'$?

\end{probl}

\begin{todo}
Multisource problem for dijkstra?
\end{todo}
\flushchapter
